---
title: "Build Consolidated Database"
author: "David Ory"
runtime: shiny
output:
  html_document:
    toc: true
    theme: cosmo
---

## Administration

#### Purpose
Aggregated, anonymous FasTrak transactions by transaction date, plaza, and tag holder zip card are provided by the FasTrak project team.  This script consolidates the year-specific data files into a standardized database for easy subsequent use in Tableau or R. 

#### Outputs
1 A full database, loaded into SQL server, for use in Tableau Server
2 A condensed database, loaded into SQL server, for use in Tableau Public
3 In-line Shiny tables and maps --> `Interactives.Rmd`

#### _TODO_
1.  Add second-half 2014 data when ready
2.  After 1, push full data to SQL
3.  After 2, rebuild full Tableau Server
4.  After 1, rebuild condensed Tableau
5.  After 5, publish condensed Tableau

## Overhead
```{r overhead}
library(knitr)
library(shiny)
library(reshape)
library(timeDate)
library(stringr)
suppressMessages(library(dplyr))
```

```{r config, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

## Year 2010 Data Read and Processing
```{r year-2010}
input <- read.table(file = "M:/Data/FasTrak/Frequency by Zipcode/2010/Year 2010 FasTrak Flat File.csv", header = TRUE, sep = ",", 
                    stringsAsFactors = FALSE)

# Standardize variable names
names(input)[names(input) == "PLAZA_NAME"] <- "Plaza"
names(input)[names(input) == "ZIP_CODE"] <- "Zipcode"

# Melt dates to rows
reshaped.2010 <- melt(input, id = c("Plaza", "Zipcode"))
reshaped.2010 <- select(reshaped.2010, Plaza, Zipcode, Date_Factor = variable, Transactions_String = value)

# clean up data types and remove zero transactions
reshaped.2010 <- reshaped.2010 %>%
  mutate(Transactions = strtoi(Transactions_String)) %>%
  filter(Transactions > 0) %>%
  mutate(Date_Temp1 = paste(sub("X","",Date_Factor))) %>%
  mutate(Date_Temp2 = paste(sub("[[:punct:]]","-",Date_Temp1))) %>%
  mutate(Date_Temp3 = paste(Date_Temp2,"10")) %>%
  mutate(Date = str_replace_all(Date_Temp3," ","-")) %>%
  select(Date, Zipcode, Plaza, Transactions)

head(reshaped.2010, n = 20)
```

## Year 2011 Data Read and Processing
```{r year-2011}
input <- read.table(file = "M:/Data/FasTrak/Frequency by Zipcode/2011/Year 2011 FasTrak Flat File.csv", header = TRUE, sep = ",", 
                    stringsAsFactors = FALSE)

# Standardize variable names
names(input)[names(input) == "PLAZA_NAME"] <- "Plaza"
names(input)[names(input) == "ZIP_CODE"] <- "Zipcode"

# Melt dates to rows
reshaped.2011 <- melt(input, id = c("Plaza", "Zipcode"))
reshaped.2011 <- select(reshaped.2011, Plaza, Zipcode, Date_Factor = variable, Transactions_String = value)

# Clean up data types and remove zero transactions
reshaped.2011 <- reshaped.2011 %>%
  mutate(Transactions = strtoi(Transactions_String)) %>%
  filter(Transactions > 0) %>%
  mutate(Date_Temp1 = paste(sub("X","",Date_Factor))) %>%
  mutate(Date_Temp2 = paste(sub("[[:punct:]]","-",Date_Temp1))) %>%
  mutate(Date_Temp3 = paste(Date_Temp2,"11")) %>%
  mutate(Date = str_replace_all(Date_Temp3," ","-")) %>%
  select(Date, Zipcode, Plaza, Transactions)

head(reshaped.2011, n = 20)
```

## Year 2012 Data Read and Processing
```{r year-2012}
input <- read.table(file = "M:/Data/FasTrak/Frequency by Zipcode/2012/Year 2012 FasTrak Flat File.csv", header = TRUE, sep = ",", 
                    stringsAsFactors = FALSE)

# Standardize variable names
names(input)[names(input) == "PLAZA"] <- "Plaza"
names(input)[names(input) == "ZIP_CODE"] <- "Zipcode"

# Melt dates to rows
reshaped.2012 <- melt(input, id = c("Plaza", "Zipcode"))
reshaped.2012 <- select(reshaped.2012, Plaza, Zipcode, Date_Factor = variable, Transactions_String = value)

# Clean up data types and remove zero transactions
reshaped.2012 <- reshaped.2012 %>%
  mutate(Transactions = strtoi(Transactions_String)) %>%
  filter(Transactions > 0) %>%
  mutate(Date_Temp1 = paste(sub("X","",Date_Factor))) %>%
  mutate(Date_Temp2 = paste(sub("[[:punct:]]","-",Date_Temp1))) %>%
  mutate(Date_Temp3 = paste(Date_Temp2,"12")) %>%
  mutate(Date = str_replace_all(Date_Temp3," ","-")) %>%
  select(Date, Zipcode, Plaza, Transactions)

head(reshaped.2012, n = 20)
```

## Year 2013 Data Reads and Processing
```{r year-2013}
# Note: data delivered in two chunks

# Data read A
input <- read.table(file = "M:/Data/FasTrak/Frequency by Zipcode/2013/Year 2013 FasTrak Flat File Jan to Jul.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)

# Standardize variable names
names(input)[names(input) == "PLAZA"] <- "Plaza"
names(input)[names(input) == "ZIP"] <- "Zipcode"

# Melt dates to rows
reshaped.2013.A <- melt(input, id = c("Plaza", "Zipcode"))
reshaped.2013.A <- select(reshaped.2013.A, Plaza, Zipcode, Date_Factor = variable, Transactions_String = value)

# Data read B
input <- read.table(file = "M:/Data/FasTrak/Frequency by Zipcode/2013/Year 2013 FasTrak Flat File Aug to Dec.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)

# Standardize variable names
names(input)[names(input) == "PLAZA"] <- "Plaza"
names(input)[names(input) == "ZIP"] <- "Zipcode"

# Melt dates to rows
reshaped.2013.B <- melt(input, id = c("Plaza", "Zipcode"))
reshaped.2013.B <- select(reshaped.2013.B, Plaza, Zipcode, Date_Factor = variable, Transactions_String = value)

# Merge A and B
reshaped.2013 <- rbind(reshaped.2013.A, reshaped.2013.B)

# Southern California Plazas included in data -- filter for Northern California plazas
reshaped.2013 <- reshaped.2013 %>%
  filter(Plaza == 'Antioch Bridge' |
           Plaza == 'Bay Bridge' |
           Plaza == 'Benicia' |
           Plaza == 'Carquinez Bridge' |
           Plaza == 'Dumbarton Bridge' |
           Plaza == 'Golden Gate Bridge' |
           Plaza == 'I-680S Calaveras' |
           Plaza == 'I-680S Mission' |
           Plaza == 'Richmond' |
           Plaza == 'San Mateo' |
           Plaza == 'SR 237 First WB' |
           Plaza == 'SR237/I880 ExpressEB')

table(reshaped.2013$Plaza)

# Clean up data types and remove zero transactions
reshaped.2013 <- reshaped.2013 %>%
  mutate(Transactions = strtoi(Transactions_String)) %>%
  filter(Transactions > 0) %>%
  mutate(Date_Temp1 = paste(sub("X","",Date_Factor))) %>%
  mutate(Date_Temp2 = paste(sub("[[:punct:]]","-",Date_Temp1))) %>%
  mutate(Date_Temp3 = paste(Date_Temp2,"13")) %>%
  mutate(Date = str_replace_all(Date_Temp3," ","-")) %>%
  select(Date, Zipcode, Plaza, Transactions)

head(reshaped.2013, n = 20)

```

## Year 2014 Data Reads and Processing
```{r year-2014}
# Note: data delivered in two chunks

# Data read A
input <- read.table(file = "M:/Data/FasTrak/Frequency by Zipcode/2014/Year 2014 FasTrak Flat File Jan to Jul.csv", header = TRUE, sep = ",",stringsAsFactors = FALSE)

# Standardize variable names
names(input)[names(input) == "EXTERN_PLAZA_ID"] <- "Plaza"
names(input)[names(input) == "ZIP_CODE"] <- "Zipcode"

# Melt dates to rows
reshaped.2014.A <- melt(input, id = c("Plaza", "Zipcode"))
reshaped.2014.A <- reshaped.2014.A %>%
  select(Plaza, Zipcode, Date_Factor = variable, Transactions_String = value) %>%
  filter(Plaza == 2 |
           Plaza == 3 |
           Plaza == 4 |
           Plaza == 5 |
           Plaza == 6 |
           Plaza == 7 |
           Plaza == 8) %>%
  mutate(Plaza = ifelse(Plaza == 2L, 'Antioch Bridge',   Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 3L, 'Richmond Bridge',  Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 4L, 'Bay Bridge',       Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 5L, 'San Mateo Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 6L, 'Dumbarton Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 7L, 'Carquinez Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 8L, 'Benicia Bridge',   Plaza))

# Data read B (forthcoming)
#input <- read.table(file = "M:/Data/FasTrak/Frequency by Zipcode/2014/Year 2014 FasTrak Flat File Aug to Dec.csv", header = TRUE, sep = ",",stringsAsFactors = FALSE)

# Standardize variable names
#names(input)[names(input) == "PLAZA"] <- "Plaza"
#names(input)[names(input) == "ZIP"] <- "Zipcode"

# Melt dates to rows
#reshaped.2014.B <- melt(input, id = c("Plaza", "Zipcode"))
#reshaped.2014.B <- select(reshaped.2014.B, Plaza, Zipcode, Date_Factor = variable, Transactions_String = value)

# Merge A and B
reshaped.2014 <- reshaped.2014.A #rbind(reshaped.2014.A, reshaped.2014.B)

# Clean up data types and remove zero transactions
reshaped.2014 <- reshaped.2014 %>%
  mutate(Transactions = strtoi(Transactions_String)) %>%
  filter(Transactions > 0) %>%
  mutate(Date_Temp1 = paste(sub("X","",Date_Factor))) %>%
  mutate(Date_Temp2 = paste(sub("[[:punct:]]","-",Date_Temp1))) %>%
  mutate(Date_Temp3 = paste(Date_Temp2,"14")) %>%
  mutate(Date = str_replace_all(Date_Temp3," ","-")) %>%
  select(Date, Zipcode, Plaza, Transactions)

head(reshaped.2014, n = 20)

```

## Combine Datasets
```{r combine-datasets}

# Bring in county via zipcode
input <- read.table(file = "reference_ca_zipcode.csv", header = TRUE, sep = ",")

zip_details <- input %>%
  select(Zipcode = ZIP, County = CntyName) %>%
  mutate(Zipcode = as.character(paste(Zipcode)))

reshaped = rbind(reshaped.2010, reshaped.2011, reshaped.2012, reshaped.2013, reshaped.2014)
reshaped = left_join(reshaped, zip_details, by = "Zipcode")

table(reshaped$County)

# Standardize plaza names
reshaped <- reshaped %>%
  mutate(Plaza = ifelse(Plaza == 'CALT/Antioch Bridge',     'Antioch Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'CALT/Bay Bridge',         'Bay Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'HYBD/Bay Bridge HOV',     'Bay Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'Benicia',                 'Benicia Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'CALT/Benicia',            'Benicia Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'CALT/Carquinez Bridge',   'Carquinez Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'HYBD/Carquinez HOV',      'Carquinez Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'CALT/Dumbarton Bridge',   'Dumbarton Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'HYBD/Dumbarton HOV',      'Dumbarton Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'CALT/Richmond',           'Richmond Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'Richmond',                'Richmond Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'CALT/San Mateo',          'San Mateo Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'San Mateo',               'San Mateo Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'HYBD/San Mateo HOV',      'San Mateo Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'GGB/Golden Gate Bridge',  'Golden Gate Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'GGBD/Golden Gate Bridge', 'Golden Gate Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'I-680/I-680S Calaveras',  'Express Lanes: I-680', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'I-680/I-680S Mission',    'Express Lanes: I-680', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'I-680S Calaveras',        'Express Lanes: I-680', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'I-680S Mission',          'Express Lanes: I-680', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'I680S/I-680S Calaveras',  'Express Lanes: I-680', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'I680S/I-680S Mission',    'Express Lanes: I-680', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'SR 237 First WB',         'Express Lanes: SR 237/I-880', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'SR 237/SR 237 First WB',  'Express Lanes: SR 237/I-880', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'SR237/I880 ExpressEB',    'Express Lanes: SR 237/I-880', Plaza))
table(reshaped$Plaza)

# Extract typical weekday data
reshaped.typical <- reshaped %>%
  mutate(Year = format(as.Date(Date, format = "%d-%b-%y"), "%Y")) %>%
  mutate(Month = format(as.Date(Date, format = "%d-%b-%y"), "%B")) %>%
  mutate(Day_of_week = weekdays(as.Date(Date, format = "%d-%b-%y"))) %>%
  filter(Month == 'March' | Month == 'April' | Month == 'May' | Month == 'September' | Month == 'October' | Month == 'November') %>%
  filter(Day_of_week == 'Tuesday' | Day_of_week == 'Wednesday' | Day_of_week == 'Thursday')

# Count the number of typical weekdays
typical.count <- reshaped.typical %>%
  select(Year, Month, Day_of_week, Date) %>%
  group_by(Year, Date) %>%
  summarise(temp = n()) %>%
  select(Year, Date) %>%
  group_by(Year) %>%
  summarise(Count_Typical_Days = n())

# Take the typical weekday average and median
typical.measures <- reshaped.typical %>%
  select(Plaza, County, Zipcode, Year, Transactions) %>%
  group_by(Year, Zipcode, Plaza, County) %>%
  summarise(Count_Days_Obs = n(), Sum_Trans_on_Days_Obs = sum(Transactions))

# Join with the typical day count to get mean
typical.measures <- left_join(typical.measures, typical.count, by = c("Year"))

# Note: some zip/plazas will have more observed days than there are typical days because two plazas (HOV and GP) are combined into one plaza
typical.measures <- typical.measures %>%
  mutate(Avg_Transactions = Sum_Trans_on_Days_Obs / Count_Typical_Days)


```

## Write data to disk
```{r data-writes}
# Full
write.csv(reshaped, file = "M:/Data/FasTrak/Frequency by Zipcode/FasTrak Transactions.csv", row.names = FALSE, quote = F)

# Typical measures (as CSV and .RData)
write.csv(typical.measures, file = "M:/Data/FasTrak/Frequency by Zipcode/Typical Weekday FasTrak Transactions Summary.csv", row.names = FALSE, quote = F)

save(typical.measures, file = "M:/Data/FasTrak/Frequency by Zipcode/Typical Weekday FasTrak Transactions Summary.Rdata")

```


