---
title: "Build Consolidated Database"
author: "David Ory"
runtime: shiny
output:
  html_document:
    toc: true
    theme: cosmo
---

## Administration

#### Purpose
Aggregated, anonymous FasTrak transactions by transaction date, plaza, and tag holder zip card are provided by the FasTrak project team.  This script consolidates the year-specific data files into a standardized database for easy subsequent use in Tableau or R. 

#### Outputs
1 A full database, loaded into SQL server, for use in Tableau Server
2 A condensed database, loaded into SQL server, for use in Tableau Public
3 In-line Shiny tables and maps --> `Interactives.Rmd`

#### _TODO_
1 Build plaza ID to plaza name cross walk for 2014
2 Add 2014 to rbind
3 Add full file to SQL (here or off-line?) when 2014
4 Rebuild (both) Tableau 

## Overhead
```{r overhead}
library(knitr)
library(shiny)
library(reshape)
library(timeDate)
library(stringr)
suppressMessages(library(dplyr))
```

```{r config, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

## Year 2010 Data Read and Processing
```{r year-2010}
input <- read.table(file = "M:/Data/FasTrak/Frequency by Zipcode/2010/Year 2010 FasTrak Flat File.csv", header = TRUE, sep = ",", 
                    stringsAsFactors = FALSE)

# Standardize variable names
names(input)[names(input) == "PLAZA_NAME"] <- "Plaza"
names(input)[names(input) == "ZIP_CODE"] <- "Zipcode"

# Melt dates to rows
reshaped.2010 <- melt(input, id = c("Plaza", "Zipcode"))
reshaped.2010 <- select(reshaped.2010, Plaza, Zipcode, Date_Factor = variable, Transactions_String = value)

# clean up data types and remove zero transactions
reshaped.2010 <- reshaped.2010 %>%
  mutate(Transactions = strtoi(Transactions_String)) %>%
  filter(Transactions > 0) %>%
  mutate(Date_Temp1 = paste(sub("X","",Date_Factor))) %>%
  mutate(Date_Temp2 = paste(sub("[[:punct:]]","-",Date_Temp1))) %>%
  mutate(Date_Temp3 = paste(Date_Temp2,"10")) %>%
  mutate(Date = str_replace_all(Date_Temp3," ","-")) %>%
  select(Date, Zipcode, Plaza, Transactions)

head(reshaped.2010, n = 20)
```

## Year 2011 Data Read and Processing
```{r year-2011}
input <- read.table(file = "M:/Data/FasTrak/Frequency by Zipcode/2011/Year 2011 FasTrak Flat File.csv", header = TRUE, sep = ",", 
                    stringsAsFactors = FALSE)

# Standardize variable names
names(input)[names(input) == "PLAZA_NAME"] <- "Plaza"
names(input)[names(input) == "ZIP_CODE"] <- "Zipcode"

# Melt dates to rows
reshaped.2011 <- melt(input, id = c("Plaza", "Zipcode"))
reshaped.2011 <- select(reshaped.2011, Plaza, Zipcode, Date_Factor = variable, Transactions_String = value)

# Clean up data types and remove zero transactions
reshaped.2011 <- reshaped.2011 %>%
  mutate(Transactions = strtoi(Transactions_String)) %>%
  filter(Transactions > 0) %>%
  mutate(Date_Temp1 = paste(sub("X","",Date_Factor))) %>%
  mutate(Date_Temp2 = paste(sub("[[:punct:]]","-",Date_Temp1))) %>%
  mutate(Date_Temp3 = paste(Date_Temp2,"11")) %>%
  mutate(Date = str_replace_all(Date_Temp3," ","-")) %>%
  select(Date, Zipcode, Plaza, Transactions)

head(reshaped.2011, n = 20)
```

## Year 2012 Data Read and Processing
```{r year-2012}
input <- read.table(file = "M:/Data/FasTrak/Frequency by Zipcode/2012/Year 2012 FasTrak Flat File.csv", header = TRUE, sep = ",", 
                    stringsAsFactors = FALSE)

# Standardize variable names
names(input)[names(input) == "PLAZA"] <- "Plaza"
names(input)[names(input) == "ZIP_CODE"] <- "Zipcode"

# Melt dates to rows
reshaped.2012 <- melt(input, id = c("Plaza", "Zipcode"))
reshaped.2012 <- select(reshaped.2012, Plaza, Zipcode, Date_Factor = variable, Transactions_String = value)

# Clean up data types and remove zero transactions
reshaped.2012 <- reshaped.2012 %>%
  mutate(Transactions = strtoi(Transactions_String)) %>%
  filter(Transactions > 0) %>%
  mutate(Date_Temp1 = paste(sub("X","",Date_Factor))) %>%
  mutate(Date_Temp2 = paste(sub("[[:punct:]]","-",Date_Temp1))) %>%
  mutate(Date_Temp3 = paste(Date_Temp2,"12")) %>%
  mutate(Date = str_replace_all(Date_Temp3," ","-")) %>%
  select(Date, Zipcode, Plaza, Transactions)

head(reshaped.2012, n = 20)
```

## Year 2013 Data Reads and Processing
```{r year-2013}
# Note: data delivered in two chunks

# Data read A
input <- read.table(file = "M:/Data/FasTrak/Frequency by Zipcode/2013/Year 2013 FasTrak Flat File Jan to Jul.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)

# Standardize variable names
names(input)[names(input) == "PLAZA"] <- "Plaza"
names(input)[names(input) == "ZIP"] <- "Zipcode"

# Melt dates to rows
reshaped.2013.A <- melt(input, id = c("Plaza", "Zipcode"))
reshaped.2013.A <- select(reshaped.2013.A, Plaza, Zipcode, Date_Factor = variable, Transactions_String = value)

# Data read B
input <- read.table(file = "M:/Data/FasTrak/Frequency by Zipcode/2013/Year 2013 FasTrak Flat File Aug to Dec.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)

# Standardize variable names
names(input)[names(input) == "PLAZA"] <- "Plaza"
names(input)[names(input) == "ZIP"] <- "Zipcode"

# Melt dates to rows
reshaped.2013.B <- melt(input, id = c("Plaza", "Zipcode"))
reshaped.2013.B <- select(reshaped.2013.B, Plaza, Zipcode, Date_Factor = variable, Transactions_String = value)

# Merge A and B
reshaped.2013 <- rbind(reshaped.2013.A, reshaped.2013.B)

# Southern California Plazas included in data -- filter for Northern California plazas
reshaped.2013 <- reshaped.2013 %>%
  filter(Plaza == 'Antioch Bridge' |
           Plaza == 'Bay Bridge' |
           Plaza == 'Benicia' |
           Plaza == 'Carquinez Bridge' |
           Plaza == 'Dumbarton Bridge' |
           Plaza == 'Golden Gate Bridge' |
           Plaza == 'I-680S Calaveras' |
           Plaza == 'I-680S Mission' |
           Plaza == 'Richmond' |
           Plaza == 'San Mateo' |
           Plaza == 'SR 237 First WB' |
           Plaza == 'SR237/I880 ExpressEB')

table(reshaped.2013$Plaza)

# Clean up data types and remove zero transactions
reshaped.2013 <- reshaped.2013 %>%
  mutate(Transactions = strtoi(Transactions_String)) %>%
  filter(Transactions > 0) %>%
  mutate(Date_Temp1 = paste(sub("X","",Date_Factor))) %>%
  mutate(Date_Temp2 = paste(sub("[[:punct:]]","-",Date_Temp1))) %>%
  mutate(Date_Temp3 = paste(Date_Temp2,"13")) %>%
  mutate(Date = str_replace_all(Date_Temp3," ","-")) %>%
  select(Date, Zipcode, Plaza, Transactions)

head(reshaped.2013, n = 20)

```

## Year 2014 Data Reads and Processing
```{r year-2014}
# Note: data delivered in two chunks

# Data read A
input <- read.table(file = "M:/Data/FasTrak/Frequency by Zipcode/2014/Year 2014 FasTrak Flat File Jan to Jul.csv", header = TRUE, sep = ",",stringsAsFactors = FALSE)

# Standardize variable names
names(input)[names(input) == "EXTERN_PLAZA_ID"] <- "Plaza"
names(input)[names(input) == "ZIP_CODE"] <- "Zipcode"

# Melt dates to rows
reshaped.2014.A <- melt(input, id = c("Plaza", "Zipcode"))
reshaped.2014.A <- select(reshaped.2014.A, Plaza, Zipcode, Date_Factor = variable, Transactions_String = value)

# Data read B (forthcoming)
#input <- read.table(file = "M:/Data/FasTrak/Frequency by Zipcode/2014/Year 2014 FasTrak Flat File Aug to Dec.csv", header = TRUE, sep = ",",stringsAsFactors = FALSE)

# Standardize variable names
#names(input)[names(input) == "PLAZA"] <- "Plaza"
#names(input)[names(input) == "ZIP"] <- "Zipcode"

# Melt dates to rows
#reshaped.2014.B <- melt(input, id = c("Plaza", "Zipcode"))
#reshaped.2014.B <- select(reshaped.2014.B, Plaza, Zipcode, Date_Factor = variable, Transactions_String = value)

# Merge A and B
reshaped.2014 <- reshaped.2014.A #rbind(reshaped.2014.A, reshaped.2014.B)

## TODO Recode Plaza IDs to names when info received
table(reshaped.2014$Plaza)

# Clean up data types and remove zero transactions
reshaped.2014 <- reshaped.2014 %>%
  mutate(Transactions = strtoi(Transactions_String)) %>%
  filter(Transactions > 0) %>%
  mutate(Date_Temp1 = paste(sub("X","",Date_Factor))) %>%
  mutate(Date_Temp2 = paste(sub("[[:punct:]]","-",Date_Temp1))) %>%
  mutate(Date_Temp3 = paste(Date_Temp2,"14")) %>%
  mutate(Date = str_replace_all(Date_Temp3," ","-")) %>%
  select(Date, Zipcode, Plaza, Transactions)

head(reshaped.2014, n = 20)

```

## Combine Datasets
```{r combine-datasets}

# Bring in county via zipcode
input <- read.table(file = "reference_ca_zipcode.csv", header = TRUE, sep = ",")

zip_details <- input %>%
  select(Zipcode = ZIP, County = CntyName) %>%
  mutate(Zipcode = as.character(paste(Zipcode)))

reshaped = rbind(reshaped.2010, reshaped.2011, reshaped.2012, reshaped.2013)
reshaped = left_join(reshaped, zip_details, by = "Zipcode")

table(reshaped$County)

# Standardize plaza names
reshaped <- reshaped %>%
  mutate(Plaza = ifelse(Plaza == 'CALT/Antioch Bridge',     'Antioch Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'CALT/Bay Bridge',         'Bay Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'HYBD/Bay Bridge HOV',     'Bay Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'Benicia',                 'Benicia Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'CALT/Benicia',            'Benicia Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'CALT/Carquinez Bridge',   'Carquinez Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'HYBD/Carquinez HOV',      'Carquinez Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'CALT/Dumbarton Bridge',   'Dumbarton Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'HYBD/Dumbarton HOV',      'Dumbarton Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'CALT/Richmond',           'Richmond Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'Richmond',                'Richmond Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'CALT/San Mateo',          'San Mateo Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'San Mateo',               'San Mateo Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'HYBD/San Mateo HOV',      'San Mateo Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'GGB/Golden Gate Bridge',  'Golden Gate Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'GGBD/Golden Gate Bridge', 'Golden Gate Bridge', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'I-680/I-680S Calaveras',  'Express Lanes: I-680', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'I-680/I-680S Mission',    'Express Lanes: I-680', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'I-680S Calaveras',        'Express Lanes: I-680', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'I-680S Mission',          'Express Lanes: I-680', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'I680S/I-680S Calaveras',  'Express Lanes: I-680', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'I680S/I-680S Mission',    'Express Lanes: I-680', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'SR 237 First WB',         'Express Lanes: SR 237/I-880', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'SR 237/SR 237 First WB',  'Express Lanes: SR 237/I-880', Plaza)) %>%
  mutate(Plaza = ifelse(Plaza == 'SR237/I880 ExpressEB',    'Express Lanes: SR 237/I-880', Plaza))
table(reshaped$Plaza)

# Extract typical weekday data
reshaped.typical <- reshaped %>%
  mutate(Year = format(as.Date(Date, format = "%d-%b-%y"), "%Y")) %>%
  mutate(Month = format(as.Date(Date, format = "%d-%b-%y"), "%B")) %>%
  mutate(Day_of_week = weekdays(as.Date(Date, format = "%d-%b-%y"))) %>%
  filter(Month == 'March' | Month == 'April' | Month == 'May' | Month == 'September' | Month == 'October' | Month == 'November') %>%
  filter(Day_of_week == 'Tuesday' | Day_of_week == 'Wednesday' | Day_of_week == 'Thursday')

# Take the typical weekday average and median
typical.measures <- reshaped.typical %>%
  select(Zipcode, Plaza, Year, Transactions, County) %>%
  group_by(Year, Zipcode, Plaza, County) %>%
  summarise(Avg_Transactions = mean(Transactions), Median_Transactions = median(Transactions), Count_Typical_Days = n())

```

## Write data to disk
```{r data-writes}
# Full
write.csv(reshaped, file = "M:/Data/FasTrak/Frequency by Zipcode/FasTrak Transactions.csv", row.names = FALSE, quote = F)

# Typical measures (as CSV and .RData)
write.csv(typical.measures, file = "M:/Data/FasTrak/Frequency by Zipcode/Typical Weekday FasTrak Transactions Summary.csv", row.names = FALSE, quote = F)

save(typical.measures, file = "M:/Data/FasTrak/Frequency by Zipcode/Typical Weekday FasTrak Transactions Summary.Rdata")

```


