---
title: "Interactives"
author: "David Ory"
runtime: shiny
output:
  html_document:
    theme: cosmo
---


#### Purpose
Aggregated, anonymous FasTrak transactions by transaction date, plaza, and tag holder zip card are provided by the FasTrak project team.  `Build Consolidated Database.Rmd` consolidates the year-specific data files into a standardized database for easy subsequent use in Tableau or R.  This script presents simple plaza shares by county. 

#### Outputs
1 In-line Shiny Tables and Charts

```{r overhead, echo=FALSE}
library(knitr)
library(shiny)
library(ggplot2)
library(scales)
library(stringr)
suppressMessages(library(dplyr))
```

```{r config, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```


## Share of Typical Weekday Bridge Transactions by County (2010 to 2014)

```{r prepare, echo=FALSE}
load("M:/Data/FasTrak/Frequency by Zipcode/Typical Weekday FasTrak Transactions Summary.Rdata")

interactive <- typical.measures %>%
  mutate(Bay_Area_County = "Outside of Bay Area")

attach(interactive)
interactive$Bay_Area_County[County == "ALAMEDA"] <- "Alameda"
interactive$Bay_Area_County[County == "CONTRA COSTA"] <- "Contra Costa"
interactive$Bay_Area_County[County == "MARIN"] <- "Marin"
interactive$Bay_Area_County[County == "NAPA"] <- "Napa"
interactive$Bay_Area_County[County == "SAN FRANCISCO"] <- "San Francisco"
interactive$Bay_Area_County[County == "SAN MATEO"] <- "San Mateo"
interactive$Bay_Area_County[County == "SANTA CLARA"] <- "Santa Clara"
interactive$Bay_Area_County[County == "SOLANO"] <- "Solano"
interactive$Bay_Area_County[County == "SONOMA"] <- "Sonoma"
interactive$Bay_Area_County[County == "SACRAMENTO"] <- "Sacramento"
interactive$Bay_Area_County[County == "STANISLAUS"] <- "Stanislaus"
interactive$Bay_Area_County[County == "SAN JOAQUIN"] <- "San Joaquin"
interactive$Bay_Area_County[is.na(County)] <- "Unknown or Outside of CA"
detach(interactive)

interactive.county.sum <- interactive %>% 
  group_by(Plaza, Bay_Area_County) %>% 
  summarise(Sum_Transactions = sum(Avg_Transactions))

interactive.plaza.sum <- interactive.county.sum %>%
  group_by(Plaza) %>%
  summarise(Plaza_Sum = sum(Sum_Transactions))

interactive.ready <- left_join(interactive.county.sum, interactive.plaza.sum, by = "Plaza")
interactive.ready <- mutate(interactive.ready, Plaza_Share = Sum_Transactions/Plaza_Sum)

```

```{r shiny_bridge_shares, echo=FALSE, cache=FALSE}
inputPanel(
  selectInput("plaza_name", label = "Select a Plaza:", choices = unique(interactive.ready$Plaza), selected = "Bay Bridge")
  )

renderPlot({
  
  data <- filter(interactive.ready, Plaza == input$plaza_name)
  
  data$county_order <- factor(data$Bay_Area_County, 
                                   c("Alameda","Contra Costa","Marin","Napa","San Francisco",
                                     "San Mateo","Santa Clara","Solano","Sonoma","Sacramento",
                                     "San Joaquin","Stanislaus","Outside of Bay Area","Unknown or Outside of CA"))
  
  plot <- ggplot(data, aes(x = Bay_Area_County, y = Plaza_Share, fill = Bay_Area_County)) + 
    geom_bar(aes(x = county_order), stat = "identity") + xlab("Bay Area County") + ylab("Share of Plaza Transactions") + 
    theme(text = element_text(size=20), axis.text.x = element_text(angle=90, vjust=1), 
          axis.title.y = element_text(vjust=2.0), legend.position="none") +
    scale_y_continuous(labels = percent)
  
  print(plot)
  
  
  }, height = 900)


```