---
title: "Build Database"
author: "David Ory"
output: 
   html_document:
      theme: cosmo
      toc: yes
---

## Administration

#### Purpose
This script consumes databases provided by the Bay Area Toll Authority of (a) toll transactions by lane by bridge by date by hour and (b) lane designations by bridge by date by hour.  These databases are consolidated and typical weekday transaction summaries are created for consumption by Tableau. 

#### Outputs
1.  A large database of typical weekday transactions by bridge by lane by date by hour
2.  A condensed database of typical weekday statistics by year 

#### TODO
1. Check with MD about a source for the lane designations

#### ISSUES

## Procedure

#### Overhead
```{r overhead, results = 'hide'}
library(knitr)
library(stringr)
library(timeDate)
library(chron)
library(lubridate)
library(reshape2)
suppressMessages(library(dplyr))
```

```{r config, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

#### Remote I/O Locations
```{r remote-io}
F_INPUT_DIRECTORY  = "M:/Data/BATA/Transactions by Lane/"
F_BOX_DIRECTORY    = "D:/files/My Box Files/Share Data/bridge-transactions/" 

F_INPUT_LANES = paste(F_INPUT_DIRECTORY, "Bridge Toll Lane Type Database.csv", sep = "")

F_OUTPUT_SUM_CSV = paste(F_BOX_DIRECTORY, "Typical Weekday Transactions Statistics.csv", sep = "")
F_OUTPUT_SUM_R   = paste(F_BOX_DIRECTORY, "Typical Weekday Transactions Statistics.Rdata", sep = "")

F_OUTPUT_TYP_CSV = paste(F_INPUT_DIRECTORY, "Typical Weekday Transactions by Lane Designation.csv", sep = "")
F_OUTPUT_TYP_R   = paste(F_INPUT_DIRECTORY, "Typical Weekday Transactions by Lane Designation.Rdata", sep = "")

# File names are bridge name + year field
bridge_names_array = c("antioch", "bay", "benicia", "carquinez", "dumbarton", "richmond", "san mateo")
year_fields_array  = c("2005 to 2013", "2014")

```

#### Plaza IDs and names
```{r plaza-id}
# Must be in same order as bridge_names array
plaza_id   = c(2, 4, 8, 7, 6, 3, 5)
plaza_name = c("Antioch Bridge", "Bay Bridge", "Benicia Bridge", "Carquinez Bridge", "Dumbarton Bridge", "Richmond Bridge", "San Mateo Bridge")

plaza.convert = data.frame(plaza_id, plaza_name)

```

#### Hour string to hour int dataframe
```{r hour-convert}
hour.convert = data.frame(hour_string = c('0000-0100','0100-0200','0200-0300','0300-0400','0400-0500',
                                          '0500-0600','0600-0700','0700-0800','0800-0900','0900-1000',
                                          '1000-1100','1100-1200','1200-1300','1300-1400','1400-1500',
                                          '1500-1600','1600-1700','1700-1800','1800-1900','1900-2000',
                                          '2000-2100','2100-2200','2200-2300','2300-2400'),
                          hour_int = c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23))
                          

```

#### Relevant holiday statics
```{r holidays}
holiday_list  <- c("USLaborDay", "USMemorialDay", "USThanksgivingDay", "USVeteransDay")
holiday_dates <- dates(as.character(holiday(2000:2020, holiday_list)), format = "Y-M-D")

```

#### Data reads
```{r data-reads}
# Bridge-specific transacitons
for (bridge_index in 1:length(bridge_names_array)) {
  for (year_index in 1:length(year_fields_array)) {
    
    input_file_name = paste(F_INPUT_DIRECTORY, bridge_names_array[bridge_index]," ", year_fields_array[year_index], ".csv", sep = "")
    input <- read.table(file = input_file_name, header = TRUE, sep = ",", stringsAsFactors = FALSE)
    
    input <- mutate(input, Plaza_ID = plaza_id[bridge_index])
    
    if(bridge_index == 1 & year_index == 1) {
      data <- input
    }
    else{
      data <- rbind(data, input)
    }
  } # year
} # bridge


# Read lane designation database
input.lanes <- read.table(file = F_INPUT_LANES, header = TRUE, sep = ",", stringsAsFactors = FALSE)

```

#### Combine and clean transaction data
```{r combine-clean}
clean.transactions <- data %>%
  select(plaza_id = Plaza_ID, lane_id = Lane.ID, date_string = Date, hour_string = Col5, transactions = Col6)

# Join with parameter databases to get hour_int and plaza_name
clean.transactions <- left_join(clean.transactions, hour.convert, by = c("hour_string"))
clean.transactions <- left_join(clean.transactions, plaza.convert, by = c("plaza_id"))

```

#### Combine transaction and lane designation data
```{r combine-lane}
clean.transactions <- clean.transactions %>%
  mutate(date = as.Date(date_string, format = "%m/%d/%Y")) %>%
  select(-date_string)

clean.lanes <- input.lanes %>%
  mutate(date = as.Date(date_string, format = "%Y-%m-%d")) %>%
  select(-date_string)

clean.combined <- left_join(clean.transactions, clean.lanes, by = c("plaza_name", "lane_id",  "date", "hour_int"))

clean.combined <- clean.combined %>%
  mutate(lane_designation = ifelse(is.na(lane_designation), 'not available', lane_designation))

head(clean.combined, n = 10)
table(clean.combined$lane_designation)

```

#### Typical weekday summaries
```{r typical-weekday-date}
clean.typical <- clean.combined %>%
  mutate(month = month(date)) %>%
  filter(month == 3 | month == 4 | month == 5 | month == 9 | month == 10 | month == 11) %>%
  select(-month) %>%
  mutate(day_of_week = weekdays(date)) %>%
  filter(day_of_week == "Tuesday" | day_of_week == "Wednesday" | day_of_week == "Thursday") %>%
  select(-day_of_week) %>%
  mutate(holiday = is.holiday(date, holiday_dates)) %>%
  filter(!holiday) %>%
  select(-holiday) %>%
  mutate(year = year(date))

# Sum over all lane designations, add summary lane designation variable
sum.over_lanes <- clean.typical %>%
  group_by(date, plaza_name, plaza_id, hour_int) %>%
  summarise(count_records = n(),
            transactions = sum(transactions)) %>%
  mutate(lane_designation = "all")

# Sum over lane designations
sum.over_desig <- clean.typical %>%
  group_by(date, plaza_name, plaza_id, hour_int, lane_designation) %>%
  summarise(count_records = n(),
            transactions = sum(transactions))

# Bind the files
sum.typical <- rbind(sum.over_lanes, sum.over_desig)

```

#### Write to disk
```{r disk-writes}
save(clean.typical, file = F_OUTPUT_TYP_R)
write.csv(clean.typical, file = F_OUTPUT_TYP_CSV, row.names = FALSE, quote = F)

save(sum.typical, file = F_OUTPUT_SUM_R)
write.csv(sum.typical, file = F_OUTPUT_SUM_CSV, row.names = FALSE, quote = F)

```
