---
title: "Build Database"
author: "David Ory"
output: 
   html_document:
      theme: cosmo
      toc: yes
---

## Administration

#### Purpose
This script consumes databases provided by the Bay Area Toll Authority of (a) toll transactions by lane by bridge by date by hour and (b) lane designations by bridge by date by hour.  These databases are consolidated and typical weekday transaction summaries are created for consumption by Tableau. 

#### Outputs
1.  A large database of transactions by bridge by lane by date by hour
2.  A condensed database of typical weekday statistics by year 

#### TODO
2.  Add 2014 data in a more systematic way to establish workflow

#### ISSUES
1.  Benicia 2007 data needs to be checked, fixed

## Procedure

#### Overhead
```{r overhead, results = 'hide'}
library(knitr)
library(stringr)
suppressMessages(library(dplyr))
library(timeDate)
library(chron)
library(lubridate)
```

```{r config, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

#### Remote I/O Locations
```{r remote-io}
F_INPUT_DIRECTORY  = "M:/Data/BATA/Transactions by Lane/"
F_OUTPUT_DIRECTORY = "D:/files/My Box Files/Share Data/bridge-transactions/" 
F_INPUT_ANT = paste(F_INPUT_DIRECTORY, "antioch 2005 to 2013.csv",   sep = "")
F_INPUT_BY1 = paste(F_INPUT_DIRECTORY, "bay 2006 to 2008.csv",       sep = "")
F_INPUT_BY2 = paste(F_INPUT_DIRECTORY, "bay 2009 to 2013.csv",       sep = "")
F_INPUT_BEN = paste(F_INPUT_DIRECTORY, "benicia 2005 to 2013.csv",   sep = "")
F_INPUT_CAR = paste(F_INPUT_DIRECTORY, "carquinez 2005 to 2013.csv", sep = "")
F_INPUT_DUM = paste(F_INPUT_DIRECTORY, "dumbarton 2005 to 2013.csv", sep = "")
F_INPUT_RCH = paste(F_INPUT_DIRECTORY, "richmond 2005 to 2013.csv",  sep = "")
F_INPUT_SAN = paste(F_INPUT_DIRECTORY, "san mateo 2005 to 2013.csv", sep = "")

F_INPUT_LANES = paste(F_INPUT_DIRECTORY, "Bridge Toll Lane Type Database.csv", sep = "")

F_OUTPUT_FULL_R   = paste(F_OUTPUT_DIRECTORY, "Transactions by Lane Designation.Rdata", sep = "")
F_OUTPUT_FULL_CSV = paste(F_OUTPUT_DIRECTORY, "Transactions by Lane Designation.csv", sep = "")
F_OUTPUT_TYP_CSV  = paste(F_OUTPUT_DIRECTORY, "Typical Weekday Transactions by Lane Designation.csv", sep = "")
F_OUTPUT_TYP_R    = paste(F_OUTPUT_DIRECTORY, "Typical Weekday Transactions by Lane Designation.Rdata", sep = "")

```

#### Plaza IDs and names
```{r plaza-id}
ANT_ID = 2
RCH_ID = 3
BAY_ID = 4
SAN_ID = 5
DUM_ID = 6
CAR_ID = 7
BEN_ID = 8

plaza.convert = data.frame(plaza_id = c(ANT_ID, RCH_ID, BAY_ID, SAN_ID, DUM_ID, CAR_ID, BEN_ID),
                           plaza_name = c("Antioch Bridge", "Richmond Bridge", "Bay Bridge", "San Mateo Bridge",
                                          "Dumbarton Bridge", "Carquinez Bridge", "Benicia Bridge"))

```

#### Hour string to hour int dataframe
```{r hour-convert}
hour.convert = data.frame(hour_string = c('0000-0100','0100-0200','0200-0300','0300-0400','0400-0500',
                                          '0500-0600','0600-0700','0700-0800','0800-0900','0900-1000',
                                          '1000-1100','1100-1200','1200-1300','1300-1400','1400-1500',
                                          '1500-1600','1600-1700','1700-1800','1800-1900','1900-2000',
                                          '2000-2100','2100-2200','2200-2300','2300-2400'),
                          hour_int = c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23))
                          

```

#### Relevant holiday statics
```{r holidays}
holiday_list  <- c("USLaborDay", "USMemorialDay", "USThanksgivingDay", "USVeteransDay")
holiday_dates <- dates(as.character(holiday(2000:2020, holiday_list)), format = "Y-M-D")

```


#### Data reads
```{r data-reads}
input.ant <- read.table(file = F_INPUT_ANT, header = TRUE, sep = ",", stringsAsFactors = FALSE)
input.by1 <- read.table(file = F_INPUT_BY1, header = TRUE, sep = ",", stringsAsFactors = FALSE)
input.by2 <- read.table(file = F_INPUT_BY2, header = TRUE, sep = ",", stringsAsFactors = FALSE)
input.ben <- read.table(file = F_INPUT_BEN, header = TRUE, sep = ",", stringsAsFactors = FALSE)
input.car <- read.table(file = F_INPUT_CAR, header = TRUE, sep = ",", stringsAsFactors = FALSE)
input.dum <- read.table(file = F_INPUT_DUM, header = TRUE, sep = ",", stringsAsFactors = FALSE)
input.rch <- read.table(file = F_INPUT_RCH, header = TRUE, sep = ",", stringsAsFactors = FALSE)
input.san <- read.table(file = F_INPUT_SAN, header = TRUE, sep = ",", stringsAsFactors = FALSE)

input.lanes <- read.table(file = F_INPUT_LANES, header = TRUE, sep = ",", stringsAsFactors = FALSE)

```

#### Combine and clean transaction data
```{r combine-clean}
# Overwrite ATCAS tests with Plaza_ID integers
input.ant <- input.ant %>%
  mutate(Plaza_ID = ANT_ID)

input.bay <- rbind(input.by1, input.by2)
input.bay <- input.bay %>%
  mutate(Plaza_ID = BAY_ID)

input.ben <- input.ben %>%
  mutate(Plaza_ID = BEN_ID)

input.car <- input.car %>%
  mutate(Plaza_ID = CAR_ID)

input.dum <- input.dum %>%
  mutate(Plaza_ID = DUM_ID) 

input.rch <- input.rch %>%
  mutate(Plaza_ID = RCH_ID)

input.san <- input.san %>%
  mutate(Plaza_ID = SAN_ID)

# Combine data and rename variables
clean.transactions <- rbind(input.ant, input.bay, input.ben, input.car, input.dum, input.rch, input.san)

clean.transactions <- clean.transactions %>%
  select(plaza_id = Plaza_ID, lane_id = Lane.ID, date_string = Date, hour_string = Col5, transactions = Col6)

# Join with parameter databases to get hour_int and plaza_name
clean.transactions <- left_join(clean.transactions, hour.convert, by = c("hour_string"))
clean.transactions <- left_join(clean.transactions, plaza.convert, by = c("plaza_id"))

```

#### Combine transaction and lane designation data
```{r combine-lane}
clean.transactions <- clean.transactions %>%
  mutate(date = as.Date(date_string, format = "%m/%d/%Y")) %>%
  select(-date_string)

clean.lanes <- input.lanes %>%
  mutate(date = as.Date(date_string, format = "%Y-%m-%d")) %>%
  select(-date_string)

clean.combined <- left_join(clean.transactions, clean.lanes, by = c("plaza_name", "lane_id",  "date", "hour_int"))

clean.combined <- clean.combined %>%
  mutate(lane_designation = ifelse(is.na(lane_designation), 'not available', lane_designation))

head(clean.combined, n = 10)
table(clean.combined$lane_designation)

```

#### Typical weekday summaries
```{r typical-weekday}
clean.typical <- clean.combined %>%
  mutate(month = month(date)) %>%
  filter(month == 3 | month == 4 | month == 5 | month == 9 | month == 10 | month == 11) %>%
  select(-month) %>%
  mutate(day_of_week = weekdays(date)) %>%
  filter(day_of_week == "Tuesday" | day_of_week == "Wednesday" | day_of_week == "Thursday") %>%
  select(-day_of_week) %>%
  mutate(holiday = is.holiday(date, holiday_dates)) %>%
  filter(!holiday) %>%
  select(-holiday) %>%
  mutate(year = year(date))

head(clean.typical, n = 10)

sum.typical <- clean.typical %>%
  group_by(year, plaza_name, plaza_id, hour_int, lane_id, lane_designation) %>%
  summarise(count_typical = n(),
            sum_transactions = sum(transactions),
            average_transactions = mean(transactions),
            minimum_transactions = min(transactions),
            maximum_transactions = max(transactions),
            std_dev_transactions = sd(transactions),
            median_transactions  = median(transactions))
  
head(sum.typical, n = 10)

# We'll eventually want to sum over the bridge lanes, so check for designations that changed mid year for which we'll get multiple records
sum.unique_lanes <- sum.typical %>%
  group_by(year, plaza_name, plaza_id, hour_int, lane_id) %>%
  summarise(unique_records = n()) %>%
  mutate(record_weight = 1 / unique_records) %>%
  select(-unique_records)

sum.typical <- left_join(sum.typical, sum.unique_lanes, by = c("year", "plaza_name", "plaza_id", "hour_int", "lane_id"))

head(sum.typical, n = 10)

```

#### Write to disk
```{r disk-writes}
# save(clean.combined, file = F_OUTPUT_FULL_R)
# write.csv(clean.combined, file = F_OUTPUT_FULL_CSV, row.names = FALSE, quote = F)

save(sum.typical, file = F_OUTPUT_TYP_R)
write.csv(sum.typical, file = F_OUTPUT_TYP_CSV, row.names = FALSE, quote = F)

```
