### Block Data

Below we go through a quick review of the difference between blocks in TomTom and TIGER. 

In summary, what we find is that the TomTom blocks throw out blocks that are in medians and roads. 

```{r, include=FALSE, echo=FALSE}
library(tigris)
options(tigris_class = "sf")
options(tigris_use_cache=TRUE)
blocks_census <- tigris::blocks("CA", counties1, class="sf", year=2010)
detach("package:tigris", unload=TRUE)

blocks_tt <- st_read("equivalencies/db.gdb",layer="tt_oa03")
```

```{r}
blocks_tt$CODE <- as.character(blocks_tt$CODE)
blocks_census$tt_code <- paste0("840",blocks_census$GEOID10)
table(blocks_census$tt_code %in% blocks_tt$CODE)
```

TomTom data are missing around 2k blocks. 

Lets look at a random sample of them on a map.

```{r}
blocks_census_not_in_tt <- blocks_census[!blocks_census$tt_code %in% blocks_tt$CODE,]

mapview(sample_n(blocks_census_not_in_tt, 500))
```

The blocks in census that are not in TomTom are those in medians and roadways. 

For the blocks that are in both census and TomTom, how different are their geometries?

Lets look at a sample to get a sense visually. 

```{r}
blocks_census_sample <- sample_n(blocks_census, 100)

blocks_tt_sample <- blocks_tt[blocks_tt$CODE %in% blocks_census_sample$tt_code,]
```

```{r}
mapview(blocks_tt_sample, col.regions="green", alpha=0.8) +
  mapview(blocks_census_sample, col.regions="blue", alpha=0.8)
```





